---
title: "Lecture 5"
output:
  github_document
---

Read the new Covid-19 data

```{r}
library(tidyverse)
dt_confirmed <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv", stringsAsFactors = FALSE)

dt_deaths <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Deaths.csv", stringsAsFactors = FALSE)

dt_recovered <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Recovered.csv", stringsAsFactors = FALSE)

```

Check whether the structure is the same

```{r}
sapply(list(dt_confirmed, dt_deaths, dt_recovered), colnames)
```
 
Convert to the long format and combine
```{r}
library(tidyr)
library(lubridate)
dt_confirmed_long <- dt_confirmed %>% pivot_longer(cols = starts_with("X"), names_to = "day", values_to = "confirmed")

dt_deaths_long <- dt_confirmed %>% pivot_longer(cols = starts_with("X"), names_to = "day", values_to = "deaths")

dt_deaths_long <- dt_deaths %>% pivot_longer(cols = starts_with("X"), names_to = "day", values_to = "deaths")

dt_recovered_long <- dt_recovered %>% pivot_longer(cols = starts_with("X"), names_to = "day", values_to = "recovered")

covid_raw <- dt_confirmed_long %>% 
    inner_join(dt_deaths_long %>% select(-Lat,-Long), by = c("Province.State", "Country.Region", "day")) %>% 
    inner_join(dt_recovered_long %>% select(-Lat,-Long), by = c("Province.State", "Country.Region", "day"))

covid <- covid_raw %>% mutate(day = mdy(paste0(gsub("X","", day),"20")))

colnames(covid) <- tolower(colnames(covid))
colnames(covid)[1:2] <- c("region","country")
covid
```

Aggregate by country

```{r}
covid_country <- covid %>% select(-lat, -long, -region) %>% group_by(country, day) %>% summarise_all(sum) %>% ungroup %>% arrange(country,day)
covid_country
```

Calculate absolute growths and percentage growths
```{r}
growth <- function(x) c(NA, diff(x)/x[-length(x)])
agrowth <- function(x)c(NA, diff(x))

covid_country <- covid_country %>% group_by(country) %>% mutate_at(c("confirmed", "deaths", "recovered"), list(diff = agrowth, growth = growth))
covid_country
```


Plot all of the countries
```{r}
ggplot(aes(x = day, y = confirmed, colour = country), data = covid_country) + geom_line()
```


Plot all of the countries one by one
```{r}
ggplot(aes(x = day, y = confirmed), data = covid_country) + geom_line() + facet_wrap(~country, scales="free_y")
```

Select only countries with  10 cases in the last 7 days

```{r}
day7 <- covid_country %>% filter(day <= max(day) & day > max(day) - 7) %>% group_by(country) %>% summarize(max_cases = max(confirmed_diff)) %>% filter( max_cases > 7)
day7 %>% arrange(-max_cases)
```

Plot the growths for these cases

```{r}
ggplot(aes(x = day, y = confirmed_diff), data = covid_country %>% filter(country %in% day7$country) %>% filter(day > max(day) - 7)) + geom_line() + facet_wrap(~country, scales="free_y")
```


Plot the deaths for selected countries

```{r}
ggplot(aes(x = day, y = deaths_diff), data = covid_country %>% filter(country %in% day7$country) %>% filter(day > max(day) - 7)) + geom_line() + facet_wrap(~country, scales="free_y")
```

Inspect death percentage variation
```{r}
pdeath <- covid_country %>% mutate(pdeath = deaths/confirmed) %>% filter(pdeath > 0) %>% 
    select(country, day, deaths, confirmed, pdeath) 
pdeath
```

Select only unique points 

```{r}
pdeath1 <- covid_country %>% select(country, confirmed, deaths) %>% unique %>% mutate(pdeath = deaths/confirmed) %>% filter(confirmed > 0) %>% filter(pdeath>0)
pdeath1
```


Explore distribution
```{r}
par(mfrow = c(1, 2))
hist(pdeath1$pdeath)
boxplot(pdeath1$pdeath)
```


Inspect outliers
```{r}
pdeath1 %>% filter(pdeath>0.2)
```

On which days it has occured?
```{r}
pdeath %>% filter(pdeath>0.2)
```

Explore distributions with outliers removed

```{r}
pdeath2 <- pdeath1 %>% filter(pdeath < 0.2)
par(mfrow = c(1, 2))
hist(pdeath2$pdeath)
boxplot(pdeath2$pdeath)
```

Do a scatter plot
```{r}
ggplot(aes(x = confirmed, y = deaths), data = pdeath2) + geom_point() + geom_smooth(method="lm")
```

```{r}
ggplot(aes(x = log(confirmed), y = log(deaths)), data = pdeath2) + geom_point(aes(colour=country)) 
```